From Questions to Queries – Enhancing the Curt System by a Theory and Implementation of Embedded WH-Questions

### 2008-09-01 thesis/paper.tex
FINAL

\usepackage{fontspec}
\usepackage[xetex,colorlinks]{hyperref}
\begin{document}
\lstset{language=Prolog,numbers=left,numberstyle=\tiny,numbersep=5pt,stepnumber=2,basicstyle=\ttfamily,identifierstyle=\textbf,keywordstyle=\underbar}
\renewcommand{\thepage}{\roman{page}}
\begin{titlepage}
  \maketitle
  \thispagestyle{empty}
  \begin{center}
    \vskip 4em
    Thesis submitted for the Degree of Bachelor of Arts\\

    \paragraph{Supervisor:} Dr. Frank Richter
    \vskip 6em
    Seminar für Sprachwissenschaft\\
    Eberhard-Karls-Universität\\
    Tübingen
  \end{center}
\end{titlepage}
\mbox{}
\newpage
\tableofcontents
\newpage
\SpruechleAufsagen{Aleksandar L. Dimitrov}
\thispagestyle{plain}
\newpage
\renewcommand{\thepage}{\arabic{page}}
\setcounter{page}{1}
\chapter{Introduction}

@@

As \acurt{} will only have to deal with answers to questions
internally, it will also only regard maximally exhaustive interpretations of
answers.

@@

\ex. \label{ex:whembed}
\a. Mia knows who snorts.\footnotesize
\b.  \Tree
[.$t$\\$know(?X_1.snort(X_1))(Mia)$ [.$T$ { $\lambda X.X(Mia)$\\Mia } ]
[.$IV$\\$\lambda X.know(?X_1.snort(X_1))(X)$ [.{$IV/\bar{t}$} 
{ $\lambda S.\lambda X.[  \lambda Y.[ know(Y)(X)](S) ]$\\knows } ]
[.{$\bar{t}$\\$?X_1.snort(X_1)$ } [.$\bar{t}/t$ ? ] [.{ $t$\\$snort(X_1)$ }
[.$T$ { $X_1$\\who } ]
[.$IV$ { $\lambda X.snort(X)$\\snorts } ] ] ] ]!\qsetw{5cm} ]\normalsize
\b. $know(?X_1.snort(X_1))(Mia)$\label{ex:formula}

Note that this example is using  the (syntactic) type system of \cite{gs:sawhq}
but different meaning postulates. Grammar rules, later explained in
\ref{sec:curtgrammar}, are
responsible for determining the correct order of functional application in the
untyped logical system \acurt{} is using.

@@

\ex. \label{ex:simpleembed}
\a. Mia knows whether Vincent snorts.  \footnotesize
\b. \Tree
[.$t$\\$know(?.snort(Vincent))(Mia)\wedge ?.snort(Vincent)$ [.$T$ { $\lambda X.X(Mia)$\\Mia } ]
[.$IV$\\$\lambda X.know(?.snort(Vincent))(X)$ [.{$IV/\bar{t}$} 
{ $\lambda S.\lambda X.[  \lambda Y.[ know(Y)(X) ](S) ]$\\knows } ]
[.{$\bar{t}$\\$?.snort(Vincent)$} [.$\bar{t}/t$ whether ] [.{ $t$\\$snort(Vincent)$ } 
[.$T$ { $\lambda X.X(Vincent)$\\Vincent } ]
[.$IV$ { $\lambda X.snort(X)$\\snorts } ] ] ] ]!\qsetw{5cm} ]
\normalsize
\b. $know(?.snort(Vincent))(Mia)$

@@

So far, it has been suggested that questions can be interpreted as partitions of
the logical space of the current discourse model. Furthermore, a means for
evaluating questions has been presented, again, with respect to the current
discourse model. Then these questions, can be embedded under certain verbs in
certain ways. But what is the assertion a sentence as \example{Mia knows who
snorts.} makes?

When treating issues like this, the link between philosophy and language becomes
most obvious. One can assume that the knowledge about the world will differ,
sometimes greatly, between different individuals. One will also have to assume
that every person has their own beliefs. In \emph{knowing} or \emph{believing},
or even \emph{imagining}, one can create an image of what it could possibly look
like.

@@

\wh-embedding verbs embed sets of propositions, and they establish a relation
between those propositions and their subject, one or more individuals. Depending
on the type of verb, it can be that this individual wonders about a certain
fact, reports it,  knows about it or believes it. In the latter two cases,
something more happens: the proposition is asserted to be a fact of the
individual's \emph{epistemic} or \emph{doxastic} worlds, which are worlds, just
like any other possible world.

@@

  An epistemic or Doxastic model is a triplet \pair{d,i,w'}, where
  \begin{itemize}
    \item i is a reference to a possible world $w = \pair{D,F,R}$ of the main
    model of the system, $\mathcal{M}$
    \item $d \in D$
    \item $w'$ is a possible world, defined the same way as before: $w' =
    \pair{D',F',R'}$
  \end{itemize}
\end{definition}

@@

\acurt{} is an experimental program. As such, it can serve as a ground for
experimentation to find out if this definition is better suited, or any other,
for representing the \example{true} state of affairs.

%\section{Treating Disjunction}

%Consider the following example dialogue with the \pn{Helpful Curt}
%system\footnote{In a real conversation with \curt, it will usually inform the
%user about the activity of external inferencing tools. This has been omitted and
%marked by \code{[…]}.}:

%\begin{verbatim}
  %> Mia or Jody dances.
  %[…]
  %Curt: OK.

  %> Who dances?
  %[…]
  %Curt: This question makes sense!

  %Curt: maybe mia
%\end{verbatim}

%Clearly, \curt{} should give a full answer in this situation, but instead it
%just guesses, and does not reply with something along the lines of \example{Either
%Mia or Jody}. Unfortunately, this result does not suffice for building a system
%that can properly treat embedded questions, since it does not reflect all
%available information about the discourse properly.

%A solution to this will again require the presence of the concept of
%\term{possible worlds}. When uttering the phrase \example{Mia or Jody dances.}
%then this can be understood as creating \emph{two possible scenarios}, one in
%which there is an individual named \example{Mia}, and this individual dances,
%and one in which there is an individual named \example{Jody} and the same holds
%for this individual. In effect, this can be translated to two possible worlds
%being present in the current discourse.

%\begin{definition}
  %Disjunction

  %For any formula $\rho$ holding a disjunction, this formula can be expressed as a set
  %of the constituents of the disjunction

  %\[ \rho = \varphi \vee \psi, \textrm{ then } \rho' = \set{\varphi, \psi}\]

  %%Every other operator or function can take a set of propositions of the type of
  %%$\rho'$ to produce a set of propositions containing \emph{all possible
  %%permutations} of $\rho'$. For every possible operator $\sem{$\cdot$}$ with arity
  %%$n$ and arguments $\sem{$\cdot$}(\set{\varphi,\psi})$

  %todo: more generic: just go from 1 to j, which is rho, then from j+1 to n

  %\begin{eqnarray*}
    %%\sem{$\cdot$}(\rho') \mapsto \set{\varphi,\psi}\\
    %%\sem{$\cdot$}(\sigma_1,\dots,\sigma_j,\rho )
    %\varphi
    %%\sigma \wedge \set{\varphi, \psi} \mapsto \set{\sigma\wedge\varphi,\sigma\wedge\psi}\\
    %%\sigma \to \set{\varphi, \psi} \mapsto \set{\sigma\to\varphi,\sigma\to\psi}\\
  %\end{eqnarray*}
%\end{definition}

\section{Inquisitive Semantics}\label{sec:altq}

The very latest research now focusses on providing an adequate logical semantics
for treating questions and even discourse within a dialogue. Groenendijk, in his
theory he calls ``Inquisitive Semantics'', argues for a completely new of of
treating dialogues (``Dialogue Management'') -- which naturally includes a
treatment of questions. His work on this topic has been prepared in \cite{g:is}
and is further developed in the notes to his course held at ESSLLI in Hamburg
(\cite{g:isdm}), just two weeks before work on this paper was to be finished.
Because of this time constraint, and because the latter paper is to date an
unfinished document, his theory could not be considered as a ground for the
implementation of \acurt. His findings, however, may very well serve as a ground
for further research on dialogue systems.

\subsubsection{The Rest}

\prol{} programs can be self-modifying, or dynamic. In this way the program is
able to change its own state and thus store information.

\subsubsection{Disjunction and Ambiguity}
Consider the following sentences:

\ex. \label{ex:disjunction} \a. Mia or Jody dances
\b. Either Butch snores or Vincent is using the chainsaw

\pn{Helpful Curt}, with its unpartitioned logical space and its lack of
the awareness of more than one (possible) world, would generate only one model
to represent the facts of \ref{ex:disjunction}. It is therefore only able to
generate \emph{one} possible answer to both questions, effectively accounting only
for one possible constituent of the disjunction, prefixing it with a `maybe' to
indicate its uncertainty.

Another example where \curt{} has to sacrifice exhaustiveness over its one-world
representation of the world is given in \ref{ex:ambig}

\ex.\label{ex:ambig} Every boxer likes a woman.

This will generally generate two readings, but only one model -- with only one
world. And further discourse will have to choose between one of the readings and
stick with the minimal model constructed by the model builder. This is not
really satisfactory and extending the dimensions of knowledge within \curt{} seems
necessary in order to treat such examples.

The previous work on \acurt{} enables it to generate a \term{choice point} -- and
moreover, not pursue only one path of it, but both at the same time, by
generating more than one world in its model.
Unfortunately, this will build up very large models over time and computations
with it might prove to be very complex and resource demanding. Adding such a
functionality to the system, however, is not a hard task and hopefully the
technique employed here can be further refined. %todo: maybe add a footnote
%about future prospects or so

We first derive know(john,all(X,imp(woman(X),love(X,john)))) from the syntax of
john knows whether every woman loves john, then we see if
all(X,imp(woman(X),love(X,john))) is valid (uninformative) and ridicule John, if
it's not valid and not consistent, assert his knowledge of the fact that it's
not true. If it's true, we assert he knows that.

For whether we can ridicule him if it's not true (inconsistent).
The MB/TP do a comparison \Disc to all(X,imp(woman(X),love(X,john)))

An important observation to be made here is that only \emph{assertive} verbs can
embed real questions. \emph{Stipulating} verbs only embed \emph{propositions} as
in \ref{ex:stipul}.
Nevertheless, both are treated in the following analysis, but with a focus on
the treatment of assertive \wh-embedding verbs.

When treating questions one will be confronted with many different
categorizations of questions. One of the first distinctions to make is the one
between \term{embedded questions} as in \ref{ex:embed} and \term{direct
questions} shown in \ref{ex:direct}.

As indicated already by the title, this paper will concern itself with the
treatment of the former and only briefly comment on the latter. The reasons for
this are numerous  %fixme I need a synonym for that
and this section is going to name the most important ones.

Early literature, such as \cite{karttunen:1977} only regards embedded
\wh-phrases, treating direct questions as semantically equivalent. To quote
Karttunen:

\ex. \a. John knows \emph{whether this sentence contains an embedded
question}.\label{ex:embed}
\b. Does this sentence pose a direct question?\label{ex:direct}

The present work will rely on \cite{gs:q} for providing the main guidelines for
the theoretic aspects and also lend from earlier works, most importantly
\cite{gs:sqpa}\footnotemark. Later on, in section \ref{sec:altq} we will also
discuss the very newest (at the time of writing) developments in question
semantics, depicted in \cite{g:is}.

\footnotetext{\cite{gs:sqpa} is actually a series of three articles
that have appeared before. Among them, we will mostly consider \cite{gs:sawhq}.
Both, the earlier paper and the later dissertation (and many other papers
originating in Amsterdam) have been kindly made
available electronically, free of charge, by the Universiteit van Amsterdam at
the following address: \url{http://dare.uva.nl}}

\chapter{Implementation} \label{sec:implementation}

Originally, \prol{} was conceptualized as a programming language for a subset of
first-order logic. This entails the predominant usage of \abbr{FOL} in section
\ref{sec:implementation}, but some higher order predicates introduced to the
language in recent years will also be put to use.

\section{Advertent Curt}

\acurt{} lends its name from the fact that it is not only aware of its own
knowledge and presuppositions, but `knows' about what individuals in its
model of the world are aware of. It is not in this sense a question
\emph{answering} system, but it may be seen as an early effort towards providing
such a system on a more sophisticated scale. While section \ref{sec:altq} shows
how direct questions can be integrated into the system with relative ease, focus
is still on providing a feasible method of implementing intensionality in what
is mostly a first-order logic computing environment.

It is also not constrained to being useful in a particular domain, like
\pn{CHAT-80} is.

\section{The Syntax of Embedded Questions}

The internal structure of the subset of interrogative sentences that will be
considered here is not a very complicated one. Since yes/no interrogatives  embed
simple propositions under a question operator only minimal modifications will
need to be made to the grammar in order to account for the latter. Constituent
interrogatives were already covered by \pn{Helpful Curt} and their syntactic
treatment does not differ in \acurt{}.

@@

\pn{Keller Storage} is already quite capable
of dealing with more complex constituents such as nested noun phrases, but it
has its limitations. In particular, it can only handle quantifier scope
ambiguities and does not take negation into account. Moreover, it overgenerates
with some quantifier scope ambiguities\footnote{\cite{blackburnbos:cl1} mention
sentences such as \ref{ex:kellersucks}, where Keller Storage produces six
instead of the desired five readings.
\ex. One criminal knows every owner of a hash bar\label{ex:kellersucks}

@@

The \code{whemb} and \code{ivtbar} categories have a \code{type} parameter, that dictates
their behaviour, although their respective type systems never cooperate with
each other. The type of the \code{whemb} category can hold the values
\code{propos}, for a proposition embedding complementizer, like \example{that},
and \code{question}, for a complementizer that turns a proposition into a yes/no
interrogative, like \example{if} and \example{whether} (listing \ref{whemb}).

@@

@@

\begin{lstlisting}[label=tbarCombine,caption={$\bar{t}$ is responsible for beta-converting the embedded sentence.},float]
combine(tbar:[app(WH,SRed)],[whemb:[WH],s:S]) :-
    betaConvert(S,SRed).

combine(tbar:[SRed],[s:S]) :-
    betaConvert(S,SRed).
\end{lstlisting}

@@

\section{Interpretation}

@@

We have seen, the \curt{} system is quite capable of treating questions this way.
Having implemented modality and possible worlds in such a fashion, let's go on
and implement some even cooler stuff (tense, aspect, negation, comparatives,
anaphora). Basing all this on a nice implementation of DRT (discussed in
\cite{kampreyle:drt}) like the one given in \cite{blackburnbos:cl2} might be
interesting. The Pronto \abbr{QA} system of \cite{prontoqa} already defines a
means of representing questions and queries against a Discourse Representation
Structure.

Burhans and Shapiro

Nelken and Francesz, give something that might not need sets of propositions

@@

\section{\code{advertentCurt.pl}}

\chapter{Appendix: Examples}

\subsection{An Example Model} \label{sec:examplemodel}

\begin{eqnarray*}
  M &=& \set{\pair{1,\pair{D_1,F_1,\varphi_1}},\pair{2,D_2,F_2,\varphi_2}} \\
  D_1 &=& \set{d_1,d_2,d_3} \\
  F_1 &=& \{f(0,mia,d_1), f(0,vincent,d_2), f(0,jody,d_3),\\
  && f(1,woman,\set{ d_1,d_3 }), f(1,man,\set{ d_2 }),
  f(2,like,\set{\pair{d_2,d_1},\pair{d_3,d_2}})\} \\
  \phi &=& \exists x,y,z [mia(x)\wedge woman(x)\wedge \wedge vincent(y)\wedge
  man(y)\wedge jody(z)\wedge woman(z) \\&&\wedge x \neq y \wedge x\neq z \wedge z
  \neq y\wedge like(x)(y)\wedge like(y)(z)]\\
  D_2 &=& 
\end{eqnarray*}

>>>>

\usepackage{fontspec}
\usepackage[xetex,colorlinks]{hyperref}
\begin{document}
\lstset{language=Prolog,numbers=left,numberstyle=\tiny,numbersep=5pt,stepnumber=2,basicstyle=\small\ttfamily,identifierstyle=\textbf,keywordstyle=\underbar}
\renewcommand{\thepage}{\roman{page}}
\begin{titlepage}
  \maketitle
  \thispagestyle{empty}
  \begin{center}
    \vskip 4em
    Thesis submitted for the Degree of Bachelor of Arts\\

    \paragraph{Supervisor:} Dr. Frank Richter
    \vskip 6em
    Seminar für Sprachwissenschaft\\
    Eberhard-Karls-Universität\\
    Tübingen
  \end{center}
\end{titlepage}
\tableofcontents
\newpage
\SpruechleAufsagen{Aleksandar L. Dimitrov}
\thispagestyle{plain}
\newpage
\renewcommand{\thepage}{\arabic{page}}
\setcounter{page}{1}
\chapter{Introduction}

@@

As \acurt{} will only have to deal with answers to questions
internally, it will also only regard maximally exhaustive interpretations of
answers, since a maximum degree of specificity is desired. While this impairs
the performance of  the system considerably, \acurt{} what never meant to be a
particularly graceful implementation, but only to provide a basic proof-of-concept.

@@

\ex. \label{ex:whembed}
\a. Mia knows who snorts.\footnotesize
\b. $know(?X_1.snort(X_1))(Mia)$\label{ex:formula}

Note that this example is using  the (syntactic) type system of \cite{gs:sawhq}
but different meaning postulates. Grammar rules, later explained in
\ref{sec:Syntax}, are
responsible for determining the correct order of functional application in the
untyped logical system \acurt{} is using.

@@

\ex. \label{ex:simpleembed}
\a. Mia knows whether Vincent snorts.  \footnotesize
\b. $know(?.snort(Vincent))(Mia)$

@@

So far, it has been suggested that questions can be interpreted as partitions of
the logical space of the current discourse model. Furthermore, a means for
evaluating questions has been presented, again, with respect to the current
discourse model. These questions can be embedded under certain verbs in a number
of possible combinations. But what is the assertion a sentence such as \example{Mia
knows who snorts.} makes?

When treating issues like this, the link between philosophy and language becomes
most obvious. One can assume that the knowledge about the world will differ,
sometimes greatly, between different individuals. One will also have to assume
that every person has their own beliefs. In \emph{knowing} or \emph{believing},
or even \emph{imagining}, one can create an image of what the world could
possibly look like.

@@

Wh-embedding verbs embed sets of propositions, and they establish a relation
between those propositions and their subject, which is typically one or more
individuals. Depending on the type of verb, it can be that this individual
wonders about a certain fact, reports it,  knows about it or believes it. In the
latter two cases, something interesting happens: the proposition is asserted to be a
fact of the individual's \emph{epistemic} or \emph{doxastic} worlds, which are
worlds, just like any other possible world, but usually only their `owner' can
see them. Others might only guess what they look like.

@@

  An epistemic or Doxastic model is a triplet \pair{d,i,w'}, where
  \begin{itemize}
    \item i is a reference to a possible world $w = \pair{D,F,R}$ of the main
    model of the system, $\mathcal{M}$
    \item $d \in D$, the individual this world belongs to
    \item $w'$ is a possible world, defined the same way as before: $w' =
    \pair{D',F',R'}$
  \end{itemize}
\end{definition}

@@

\chapter{Implementation} \label{sec:implementation}
%\section{Advertent Curt}

%\acurt{} lends its name from the fact that it is not only aware of its own
%knowledge, but `knows' about what individuals in its
%model of the world are aware of. It is not in this sense a question
%\emph{answering} system, but it may be seen as an early effort towards providing
%such a system on a more sophisticated scale.

With the theoretical background of the previous chapter in mind, the discussion
can now turn to the implementation of some of the more interesting concepts
presented. In particular, this chapter provides reference implementations of a
grammar that is able to parse simple embedded questions in section
\ref{sec:Syntax}, and then goes on to explain \curt's refaced discourse model in
section \ref{sec:indices}. Finally, section \ref{sec:inference} explains how the
inferencing system deals with the new extensions.

\section{The Syntax of Embedded Questions}\label{sec:Syntax}

The internal structure of the subset of interrogative sentences that is
considered here is not a very complicated one. Since yes/no interrogatives  embed
simple propositions under a question operator only minimal modifications will
need to be made to the grammar in order to account for the latter. Constituent
interrogatives were already covered by \pn{Helpful Curt} and their syntactic
treatment does not differ in \acurt{}.

@@

\pn{Keller Storage} is already quite capable of dealing with more complex
constituents such as nested noun phrases, but it has its limitations. In
particular, it can only handle \emph{quantifier} scope ambiguities and does not
take other ambiguities, such as negation into account. Moreover, it
overgenerates with some quantifier scope
ambiguities\footnote{\cite{blackburnbos:cl1} mention sentences such as
\ref{ex:kellersucks}, where Keller Storage produces six instead of the desired
five readings.  \ex. One criminal knows every owner of a hash
bar\label{ex:kellersucks}

@@

@@

The \code{whemb} and \code{ivtbar} categories have a \code{type} parameter that dictates
their behaviour, although their respective type systems never cooperate with
each other. The type of the \code{whemb} category can hold the values
\code{propos}, for a proposition embedding complementizer, like \example{that},
and \code{question}, for a complementizer that turns a proposition into a yes/no
interrogative, like \example{if} and \example{whether} (listing \ref{whemb}).

@@

\begin{lstlisting}[label=tbarCombine,caption={$\bar{t}$ is responsible for beta-converting the embedded sentence.},float]
combine(tbar:[app(WH,SRed)],[whemb:[WH],s:S]) :-
    betaConvert(S,SRed).
combine(tbar:[SRed],[s:S]) :-
    betaConvert(S,SRed).
\end{lstlisting}

@@

\prol{} programs can be self-modifying. In this way the program is able to
change its own state and thus store information. The \lstinline!assert/1!,
\lstinline+abolish/1+ and \lstinline!retract/1! goals are typically used to access a
program's logic directly. \curt{} uses these goals to store information about
the current discourse, models and the dialogue's history.

Furthermore, the internal representations of models and readings have been
folded into one entity. Every world is assigned the reading that
originally generated it via the model builder that was used during inferencing
tasks.

A slightly refurbished notion of discourse has been installed. It allows
\acurt{} to perform computations on arbitrarily large numbers of possible worlds
without losing information by having to decide on one reading. Thus the readings
of ambiguous sentences such as \example{Every boxer likes a woman} can be
pursued in parallel.  As soon as a new input sentence conflicts with one
possible world, this world is \example{dropped}, with computation continuing on
the other worlds. Listing \ref{interpret} shows how this works for one
particular world and one particular reading. The \code{interpretReadings/3}
predicate is responsible for assigning subtasks to \code{interpret/3}, and can
be looked up in the appendix. The \code{check/3} predicate is a trivial abstraction
over the original calls to the inferencing engines and is  also available in the
appendix.

\begin{lstlisting}[label=interpret,caption={The interpretation code for one
  particular world and reading},float]
interpret((Index,Old),New,World) :-
  (
    beAdvertent((Index,Old),New), !
    , World = (Index,Old)
  ;
    getKnowledge(Old,New,BK,Reading) , (
      check(and(BK,New),'consistency',BBModel), !  , (
        check(and(BK,not(New)),'informativity',_), !
        , BBModel = model(D,F)
        , World = (Index,world(D,F,Reading))
      ;
        format('~nFound uninformative reading. Dropping reading.',[])
        , World = (Index,Old)
      )
    ;
      format('~nFound inconsistency. Dropping world.',[])
      , World = [])) .
\end{lstlisting}

\subsection{Epistemic and Doxastic Alternatives}\label{sec:inference}

While the original
implementation provided only one set of models, using \lstinline!models/1!, this
implementation introduces two
new dynamic goals, \lstinline!epistemic/1! and \lstinline!doxastic/1!. They
serve the obvious purpose of accumulating what \curt{} knows about other
individual's knowledge. Listing \ref{getEpistemicBG} shows the code for the
retrieval of epistemic backgrounds. The one for doxastic backgrounds is very
similar and omitted here. The goal fails if it cannot unify with any particular
background given its instantiations, which are the name of the entity in
question (\code{X}) and the index of a possible world in the main curt storage
(\code{I}).

\begin{lstlisting}[label=getEpistemicBG,caption={A Predicate to Retrieve Epistemic
  Backgrounds},float]
getEpistemicBG(X,I,world(D,F,R)) :-
  epistemic(E), !
  , delete(E,(X,I,world(D,F,R)),Rest)
  , \+ Rest = E
  , retract(epistemic(_)), !
  , assert(epistemic(Rest)) .
\end{lstlisting}

\subsection{Interpretation of \wh-embedded Verbs}

The heart of the reasoning of \acurt{} lies within the predicate
\code{beAdvetent/3}. Listing \ref{beAdvertent} reproduces one of its clauses,
this one specifically dealing with answers to constituent interrogatives. First,
\code{findBuddies/2} returns a list of all individuals in the current discourse.
Then \code{formulate/4} is called on each name for the body and domain of the
question. This predicate instantiates an unsaturated formula with a name it is
given, to produce, e.g. $boxer(butch)$ from $butch$ and $boxer(X)$. It is
important to note that to date, \acurt{} can only deal with named individuals,
and not with quantifications over individuals, such as $\exists x[boxer(x)]$.

\begin{lstlisting}[label=beAdvertent,caption={The Core of Advertent Curt Dealing
  with Constituent Interrogatives},float]
beAdvertent((_Index,world(_D,F,R)),que(X,Domain,Body),Answer) :-
  findBuddies(F,Buddies)
  , maplist(or:formulate(X,and(Domain,Body)),Buddies, Formulae)
  , partition(Formulae,Parted)
  , checkPartitions(R,Parted,Answer) .
\end{lstlisting}

Next, the combinatorial bomb is armed with \code{partition/2}, which generates
all possible logical partitions of the domain and body of the question with
respect to the discourse referents it is given. Afterwards,
\code{checkPartitions/3} is called and ignites the inferencing tools by calling
them once for each partition, unified with the current discourse of the world
(held in the variable \code{R}) to find the one most specific partition, which
serves as the only true answer to the question.
The code for the helper predicates is again available in the appendix. Other,
more simple clauses of \code{beAdvertent/3} deal with embedded propositions and
yes/no interrogatives and are not reproduced here.

Listing \ref{beA2} shows how \acurt{} handles embedded interrogatives. It first
calls the predicate shown above, then retrieves the individual's knowledge
background and takes that as a basis for further computation. First, the
background knowledge is computed, then the resulting formula is checked by the
inferencing tools -- this time with the individual's epistemic background. This
ensures that an individual cannot hold contradictory beliefs. The check also
produces a model, which is then passed on to the epistemic background of
\code{X}.

\begin{lstlisting}[label=beA2,caption={Treatment of Embedded Interrogatives of
  Knowledge},float]
beAdvertent((Index,world(D,F,Background)),knowledge(X,que(Y,Domain,Body))) :-
	! , beAdvertent((Index,world(D,F,Background)),que(Y,Domain,Body),Answer)
	, (
		getEpistemicBG(Y,_,world(_,_,EBG))
		, !  , Q = and(Answer,EBG)
	;
		Q = Answer
	)
	, backgroundKnowledge(Q,BK)
	, check(and(Q,BK),'preparing world: consistency', model(D2,F2))
	, World = (X, Index,world(D2,F2,Q))
	, addEpistemic(World) .
\end{lstlisting}

@@

This thesis presented a method of approaching question answering in a
light-weight prototype of a dialog system. As such, not all results are
satisfactory, and in order to cover the semantics of at least some basic human
conversation, \acurt{} will have to grow considerably. Yet it is debatable
whether the \curt{} system can serve as a ground to a full-fledged dialogue
system, since it was only planned an designed as a teaching system for
\cite{blackburnbos:cl1}. It would have to be modified at many places, provided
with a more exhaustive ontology, and better parser. \curt{} uses nothing more
than the \prol{} in-built \abbr{DCG} parser, which was not designed to cover a
wide range of natural language syntax.

\acurt{} can deal with basic dialogues such as the following:
\begin{quote}
  Mia and Vincent snort. Butch does not snort. Butch is not Vincent. Butch is
  not Mia.

  Who snorts? Vincent knows who snorts. Mia knows if Butch snorts.
\end{quote}

In this scenario, \curt{} will correctly answer that Mia and Vincent snort, but
not Butch. It will also acknowledge that Vincent knows about this fact. Finally,
it will also record that Mia knows that Butch does not snort. Note that when
using its ontology, the system will not need the information of the third and
fourth sentence. The sentences will then still show how \curt{} can distinguish
between informative and uninformative readings even with respect to its
background knowledge.

The extension of the system could still use some improvements to address its
present poor usability. Currently, \acurt{} will believe everything it is told,
and when it finds a contradiction, it will just eliminate the offending worlds
from its storage. Thus, when the user inadvertently enters a contradiction,
\acurt{} will end up in a state of ignorance.

Furthermore, while \curt{} does \emph{acknowledge} the epistemic and doxastic
background of the individuals it keeps track of, it does so far not do anything
useful with them, other than display them. However, it should not pose a
significant challenge to implement a functionality that would be able to answer
a question such as \example{What does Vincent believe?}

A fully-capable dialogue system would have to encompass a much wider range of
functionality. At least a rudimentary notion of tenses and events, and a solid way of
treating anaphoric references seem to be a bare minimum of what such a system
should be capable of when talking to a human. But as it stands, \curt{} is not
even able to cope with negation properly. For example, it cannot interpret a
sentence such as \example{Vincent does not know who snorts.} And what is this
sentence supposed to mean? This could be an interesting future topic in the
research of embedded questions.

Discourse Representation Theory, as introduced in
\cite{kampreyle:drt} could probably solve some of the issues with \curt{}. It
provides a solid treatment of anaphoric references, is directly translatable
into \abbr{FOL} and Kamp \& Reyle also propose methods of dealing with plurals,
tense and negation in \abbr{DRT}.
\cite{blackburnbos:cl2} presents an approach to implementing \abbr{DRT} in
\prol{}. Furthermore, the \pn{Pronto QA System} of \cite{prontoqa} already uses
\abbr{DRT} for its inferencing tasks, and so does the implementation proposed by
\cite{bosgabdsil}.

\subsection{Inquisitive Semantics}\label{sec:altq}

The very latest research now focusses on providing an adequate logical semantics
for treating questions and even discourse within a dialogue. Groenendijk, in his
theory he calls ``Inquisitive Semantics'', argues for a completely new way of
treating dialogues (``Dialogue Management'') -- which naturally includes a
treatment of questions. His work on this topic has been prepared in \cite{g:is}
and is further developed in the notes to his course held at ESSLLI in Hamburg
(\cite{g:isdm}), just two weeks before work on this paper was to be finished.
Because of this time constraint, and because the latter paper is to date an
unfinished document, his theory could not be considered as a ground for the
implementation of \acurt. His findings, however, may very well serve as a ground
for further research on dialogue systems.

%Burhans and Shapiro

%Nelken and Francesz, give something that might not need sets of propositions

@@

\lstset{basicstyle=\tiny\ttfamily}

\section{Curt and Helpers}
\subsection{\code{advertentCurt.pl}}

\lstinputlisting{../src/advertentCurt.pl}

\subsection{\code{or.pl}}

\lstinputlisting{../src/or.pl}
\section{Grammar}
\subsection{\code{semLexStorage.pl}}
\lstinputlisting{../src/semLexStorage.pl}

\subsection{\code{semRulesKeller.pl}}
\lstinputlisting{../src/semRulesKeller.pl}

\subsection{\code{englishGrammar.pl}}
\lstinputlisting{../src/englishGrammar.pl}

\subsection{\code{englishLexicon.pl}}
\lstinputlisting{../src/englishLexicon.pl}
